Social Participation Ontology: community documentation, enhancements and use examples

### 2019-07-05 plosOne/plosArXiv.tex
feita revisao completa do artigo

             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\owli}{O\textsc{wl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urii}{U\textsc{ri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.br}
\newcommand{\ontop}{\textsc{o}n\textsc{t}op}
\newcommand{\quest}{\textsc{q}uest}
\newcommand{\webprotege}{\textsc{w}ebprotege}
\newcommand{\obda}{\textsc{obda}}
\newcommand{\pnud}{\textsc{undp}}
\newcommand{\onu}{\textsc{un}}
\newcommand{\vbs}{\textsc{vbs}}
\newcommand{\lod}{\textsc{lod}}
\newcommand{\corais}{\textsc{c}orais}
\newcommand{\serpro}{\textsc{s}erpro}
\newcommand{\python}{\textsc{p}ython}
\newcommand{\protege}{\textsc{p}rot\`eg\`e}
% triplification of data

@@

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Social Participation Ontology: community documentation, enhancements and use examples}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Renato Fabbri\textsuperscript{1},
Henrique Parra Parra Filho\textsuperscript{2},
Rodrigo Bandeira de Luna\textsuperscript{2},
Ricardo Augusto Poppi Martins\textsuperscript{3},
Flor Karina Mamani Amanqui\textsuperscript{4},
Dilvan de Abreu Moreira\textsuperscript{4}
Osvaldo Novais de Oliveira Junior\textsuperscript{1},
%\textsuperscript{3,*},
%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
%\bf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\bf{1} S\~ao Carlos Institute of Physics, University of S\~ao Paulo, CP 369, 13560-970, S\~ao Carlos, SP, Brazil, \url{fabbri@usp.br}
\\
\bf{2} Cidade Democr\'atica, Instituto Seva, S\~ao Paulo, SP, Brazil
\\
\bf{3} National Secretariat of Social Participation, General Secretariat of the Republic, Bras\'ilia, Federal District, Brazil
\\
\bf{4}  Institute of Mathematical and Computer Sciences, University of São Paulo, S\~ao Carlos, SP, Brazil
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Participatory democracy advances in virtually all governments and especially in South America
which exhibits a mixed culture and social predisposition.
This article presents the ``Social Participation Ontology'' 
(OPS from the Brazilian name \emph{Ontologia de Participação Social})
implemented in compliance with the Web Ontology Language standard (OWL)
for fostering social participation, specially in virtual platforms.
The entities and links of OPS were defined based on an extensive collaboration of specialists.
It is shown that OPS is instrumental for information retrieval from the contents of the portal,
both in terms of the actors (at various levels) as well as mechanisms and activities.
Significantly, OPS is linked to other OWL ontologies as an upper ontology
and via FOAF and BFO as higher upper ontologies,
which yields sound organization and access of knowledge and data.
In order to illustrate the usefulness of OPS, we present results on ontological expansion
and integration with other ontologies and data.
Ongoing work involves further adoption of OPS by the official
Brazilian federal portal for social participation and NGO s,
and further linkage to other ontologies for social participation.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author Summary}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
%
% \linenumbers

\section{Introduction}\label{sec:into}
Easy access to social media is reshaping citizen participation in government affairs~\cite{socMed}. 
Information and communication technologies (\ict s) have exhibited such an impact on the way individuals interact
that it is giving birth to new organizational methods in social movements.
These changes can be observed, for example, in the 2010 Arab Spring and the 2013 Brazilian protests.
These events gathered millions of people and, although recent,
have shown direct and strong impact in governments and new laws,
and the forecast is an intensification of the process~\cite{digRev1,digRev2,digRev3}.
Concomitantly, electronic government initiatives are flourishing,
favored mainly by the ubiquity of Internet technologies (e.g. \html\ 5, \node, open source browsers)
and by the need for renewal of representative democracy practices. 
These initiatives have taken place in various platforms, including usual social networks (e.g. \facebook, \twitter)
and dedicated instances created by both government and civil society parties~\cite{socMed,pita2010arquitetura,barros2010alem,knowledge}.
A natural challenge arises: how to link information produced into an unified knowledge base.
This is being addressed, at the technology level, by semantic web developments.
Endorsed by World Wide Web Consortium (\wwwc), current semantic web technologies include~\cite{Sem1}:
\begin{itemize}
    \item reasoning by means of ontological specifications;
    \item linking data from different sources (e.g. databases);
    \item organization of domain knowledge for coherent consideration.
\end{itemize}
\noindent Key among these technologies, ontologies are considered one of the pillars of the semantic web.
An ontology is usually defined as a formal specification of a shared conceptualization~\cite{gruber}.
They give meaning to data and are useful for datasets available on the web 
to make them automatically retrievable and linkable with other datasets.
The \wwwc\ created the Web Ontology Language (\owl) as a standard to represent ontologies in the web.
The second version of the language is called \owl\ 2 and offers greater expressive power~\cite{owl2}.

In this context, to describe and give meaning to social participation,
the ``Common Vocabulary of Social Participation''
(\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social})
was proposed as a joint effort of Latin America academic, civil and governmental groups~\cite{corais}.
Although started in 2012, a recent initiative for ontological developments,
it already yielded relevant material, including a public preliminary \owl\ ontology with a concise taxonomy.
Also important are the reference documents reporting results from a first working phase, from July to December, 2012. 
As stated by the community, \vcps\ was propelled by three goals:
1) to ease adoption of the vocabulary;
2) to stimulate the creation of public tools to understand, visualize and summarize how participation is happening;
3) to meet the need of participative initiatives to open and link their data. 
It is important to notice that \vcps,
an ontology, was called a \emph{vocabulary} both to ease understanding of the general public and because it started as a vocabulary.
The present article presents the ``Ontology of Social Participation'' 
(\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}),
based on \vcps, in which the term \emph{vocabulary} was substituted by the term \emph{ontology} for the following reasons:
\begin{itemize}
    \item The usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology
    (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \opsi\ is, in fact, an ontology, with a vocabulary,
    a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for academic acceptance and further formal adoption by government instances,
    such as the Brazilian Federal Portal of Open Data~\cite{dadosGov} and the Brazilian Federal Portal of Social Participation~\cite{participa}.
\end{itemize}

\noindent Also, the term \emph{common} was dropped when \ops\ was conceived,
as the term is redundant for an \owl\ ontology.
The \vcps\ presented other difficulties, such as missing classes,
incorrect \uri\ specifications (containing spaces), some logical flaws,
and unnecessary out-of-standards restrictions.
This were all solved within \ops\ (to the extent authors were able to, of course).

Next section presents \ops\ and modifications made from \vcps\ to \ops: class and property names and labels,
class restrictions and property axioms.
Section~\ref{ospUtil} is dedicated to \ops\ usage: dereferencing, \sparql\ queries,
a toy \ops\ expansion, discursive example of usage, and use cases from government,
civil society and academic parties.
Concluding remarks are stated with future work, in Section~\ref{conc}. 
The Appendixes hold directions for the scripts that are used to obtain \ops,
notes on restrictions that were in \vcps\ but withdrawn from \ops\, and an inspection of the first-hand
documentation about \vcps\ (reference textual documents, images, \owl\ code, blog posts, discussions and \etherpad s).

\section{OPS: the Ontology of Social Participation}\label{exp}
\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (\ops).
    Arrows with white heads indicate ``is a'' relations (subclass points to superclass).
    Arrows with black heads indicate property relations from subject to object .
    All property relations yield existential restrictions,
    with the exception of the ``has characteristic'' property,
    that does not yield restriction.
    Upper ontologies \bfo\ and \foaf\ are asserted through color (\bfo) and dashed boxes (\foaf).}
    \label{fig:v1}
\end{figure*}
This section makes considerations about \ops\ label standardization and implemented classes,
properties and restrictions.
Features present in Figure~\ref{fig:v1}, but not present in \vcps,
are fully described in Section~\ref{impl}.
Examples of usage are addressed in Section~\ref{ospUtil}.

\subsection{Standardization and implemented features}\label{impl}
Without explicit criteria,
\vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}.
\opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations.
    This is important to coalesce interested parties: the Brazilian Federal Social Participation Portal\cite{participa},
    the Brazilian government repository of vocabularies and ontologies~\cite{vocab},
    academic groups,  \ngo s, and non-organized civil society.
    \item Derived \uri s, when reached via \http, can be redirected to where current documentation is held,
    as it is hosted by \url{http://purl.org}.
\end{itemize}
Labels in the languages of interest should be written in label fields.
Even so, we advocate that \ops\ class names should be friendly to users,
bearing the attention not to take the class name as the label or as a meaning restriction.
For standardization, all classes are written in CamelCase~\cite{cc} in plain English to ease internationalization,
adoption and maturation.
Labels are written in English, Portuguese and Spanish.
Accordingly, class names changed from \vcps\ to \ops\,
received labels (\texttt{rdfs:label}) in each language
and a textual short explanation (\texttt{rdfs:comment}) in English. 
Table~\ref{ospClasses} exposes all classes is \ops.

\opsi\ property names fit headlessCamelCase~\cite{cc} format,
are readable in English (to ease internationalization, adoption and maturation),
and some of them have defined domains and ranges.
Table~\ref{ospProps} is dedicated to \ops\ properties, with labels in English, Portuguese and Spanish.

In the first versions of \ops, all properties yielded existential restrictions,
except {\tt ops:hasParticipationCharacteristic}.
Although such efforts were aimed at handling a better defined \ops,
further developments and discussions revealed that these restrictions made \ops\ rigid,
a bit more complicated, and was not of much help,
at least in this stage of \ops\ development and adoption.
The result is that all restrictions were removed.
Appendix~\ref{ap:restr} and Table~\ref{ospRestr} exposes all dropped restrictions from \vcps\ to \ops.

An inspection of \vcps\ reveals a contradiction present in Figure~\ref{fig:diaorig}:
a role cannot execute, initiate or support a cause.
These are done by the social actor.
Therefore, to maintain \vcps\ directions and obtain a sound ontology,
{\tt ops:Role} was removed from \ops.
The result can be seen directly in Figures~\ref{fig:v1} and~\ref{fig:owlCC}.

@@

A comparison of the \vcps\ \owl\ code~\cite{owlCCPtg} 
with the diagram in Figure~\ref{fig:diaorig}, which reflects official \vcps\ documentation,
revealed that a class, two properties and three restrictions were not implemented in \vcps.
These were fully implemented in \ops\ before all restrictions were removed.
These are the missing class and properties 
(restrictions missing in \vcps\ and implemented in preliminary \ops\ versions are exposed in Appendix~\ref{ap:restr}):
\begin{itemize}
    \item Class: {\tt    ops:ParticipationCharacteristic}.
    \item Property: {\tt ops:hasRole}.
    \item Property: {\tt ops:composesSolution}.
\end{itemize}

\opsi\ is available online~\cite{owlOSP}.
To ease navigation of the ontology by interested parties,
it is also available in the Webprotege interface~\cite{owlOSPwp}.
The diagram of \ops ' taxonomic structure is exposed in Figure~\ref{fig:owlCC}.

Upper ontologies usage with \ops\ is under development and should receive a dedicated article,
as possibilities should be inspected carefully.
Pertinent and already used as upper ontologies for \ops\ are \foaf~\cite{foaf}
(for linking and describing people and things they do) and \bfo~\cite{bfo}
(``designed for use in supporting information retrieval, analysis and integration in scientific and other domains''
as stated on their documentation).
Properties were not related to upper ontologies as reasonable relations are still being searched for.
Upper ontologies classes related to each \ops\ class are also exposed in Table~\ref{ospClasses}.

Figure~\ref{fig:v1} is a complete diagram of current \ops:
classes, properties and relations to \foaf\ and \bfo.
Actually, Figure~\ref{fig:v1} is more informative than the \ops\ \owl\ code,
as restrictions were removed and not all properties have defined domain and range.
Therefore, the diagram is an important source of relations envisioned by \ops\ creators.

@@

\section{OPS usefulness}\label{ospUtil}
\opsi\ is meant to be useful in at least two ways.
First, as a systematization of what is social participation to Latin America groups,
as conceived by \vcps.
Second, as a mean to ease linked data,
and enable integration of various instances for social participation.
An indicative of this pertinence is \opa, \ocd, \ontologiaa, and \obs,
ontologies that already uses \ops\ as upper ontology~\cite{pnud5}.

\opsi\ usage might not be obvious at first.
How is data linked? How is field knowledge organized? Why and by whom?
Core principles of \ops\ utility can be understood by the following observations:
\begin{itemize}
    \item Different participation instances have social actors,
    actions being developed, organizations involved, problems being tackled, etc.
    These can yield one consistent database by means of \ops\ usage.
    \item One can understand and share the mutually exclusive nature of being
    a paid or a voluntary contributor by observing the expanded version of \ops\ (see Section~\ref{downwards}).
    Also, noticing the fact that a mob can be very big or not,
    and that it can be convoked or not by a network,
    can make the field conceptualization more clear for a newcomer
    or ease discussions and problematization for senior researchers or politicians.
    \item Other fields of human knowledge and practice also have agents, problems and so forth.
    These can be linked to participatory data and ontologies by means of \ops.
    \item \opsi\ carry a conceptualization that is the product of contribution of many specialists.
    In other words, put aside the linked data aspects,
    it is useful as an organization of the knowledge related to the social participation field.
\end{itemize}

>>>>

             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\owli}{O\textsc{wl}}
\newcommand{\sparql}{\textsc{sparql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urii}{U\textsc{ri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.\textsc{br}}
\newcommand{\ontop}{\textsc{o}n\textsc{t}op}
\newcommand{\quest}{\textsc{q}uest}
\newcommand{\webprotege}{\textsc{w}ebprotege}
\newcommand{\obda}{\textsc{obda}}
\newcommand{\pnud}{\textsc{undp}}
\newcommand{\onu}{\textsc{un}}
\newcommand{\vbs}{\textsc{vbs}}
\newcommand{\lod}{\textsc{lod}}
\newcommand{\corais}{\textsc{c}orais}
\newcommand{\serpro}{\textsc{s}erpro}
\newcommand{\python}{\textsc{p}ython}
\newcommand{\protege}{\textsc{p}rot\`eg\`e}
% triplification of data

@@

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Social Participation Ontology: community documentation, enhancements and use examples}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Renato Fabbri\textsuperscript{1},
Henrique Parra Parra Filho\textsuperscript{2},
Rodrigo Bandeira de Luna\textsuperscript{2},
Ricardo Augusto Poppi Martins\textsuperscript{3},
Flor Karina Mamani Amanqui\textsuperscript{1},
Dilvan de Abreu Moreira\textsuperscript{1}
Osvaldo Novais de Oliveira Junior\textsuperscript{4},
%\textsuperscript{3,*},
%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
%\bf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\bf{1}  Institute of Mathematical and Computer Sciences, University of São Paulo, S\~ao Carlos, SP, Brazil
\\
\bf{2} Cidade Democrática, Instituto Seva, São Paulo, SP, Brazil
\\
\bf{3} National Secretariat of Social Participation, General Secretariat of the Republic, Brasília, Federal District, Brazil
\\
\bf{4} S\~ao Carlos Institute of Physics, University of São Paulo, CP 369, 13560-970, São Carlos, SP, Brazil, \url{renato.fabbri@gmail.com}
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Participatory democracy advances in virtually all governments and especially in South America
which exhibits a mixed culture and social predisposition.
This article presents the ``Social Participation Ontology'' 
(OPS from the Brazilian name \emph{Ontologia de Participação Social}),
implemented in compliance with the Web Ontology Language (OWL) standard,
for fostering social participation, specially in virtual platforms.
The entities and links of OPS were defined based on an extensive collaboration of specialists
and socially engaged agents, a characteristic not found, by the authors,
in other social participation-related ontologies.
OPS is linked to other OWL ontologies as an upper ontology
and via FOAF and BFO as higher upper ontologies,
which yields sound organization and access of knowledge and data.
It is shown that OPS is instrumental for retrieving contents of web portals,
both in terms of actors and of mechanisms and activities.
In order to further illustrate the usefulness of OPS, we present results on ontological expansion
and integration with other ontologies and data.
Ongoing work involves further adoption of OPS by
portals maintained by State agencies and NGOs,
and further linkage to other ontologies for social participation.

% keywords: social participation, semantic web, electronic democracy, electronic participation, OWL ontology
% \linenumbers

\section{Introduction}\label{sec:into}
Easy access to social media is reshaping citizen participation in government affairs~\cite{socMed}. 
Information and communication technologies (\ict s) have exhibited such an impact on the way individuals interact
that it is giving birth to new organizational methods in social movements.
These changes can be observed, for example, in the 2010 Arab Spring, the 2013 Brazilian protests,
and the 2018 USA March for Our Lives.
These events gathered millions of people and
have shown direct and strong impact in governments and new laws,
and the forecast is an intensification of the process~\cite{digRev1,digRev3,digRev4}.
Concomitantly, electronic government initiatives are flourishing,
favored mainly by the ubiquity of Internet technologies (e.g. \html\ 5, \node, open source browsers)
and by the need for renewal of representative democracy practices. 
These initiatives have taken place in various platforms, including usual social networks (e.g. \facebook, \twitter)
and specialized instances created by both government and civil society parties~\cite{socMed,pita2010arquitetura,barros2010alem,knowledge}.
A natural challenge arises: how to link information produced into an unified knowledge base.
This is being addressed, at the technology level, by semantic web developments.
Endorsed by the World Wide Web Consortium (\wwwc), current semantic web technologies include~\cite{Sem1}:
\begin{itemize}
    \item reasoning by means of ontological specifications;
    \item linking data from different sources (e.g. datasets);
    \item organization of domain knowledge for coherent consideration.
\end{itemize}
\noindent Key among these technologies, ontologies are considered one of the pillars of the semantic web.
An ontology is usually defined as a formal specification of a shared conceptualization~\cite{gruber}.
They give meaning to data and are useful for datasets available on the web 
to make them automatically retrievable and linkable with other datasets.
The \wwwc\ lead the creation of the Web Ontology Language (\owl) as a standard to represent ontologies on the web~\cite{owl2}.

In this context, to describe and attribute formalized meaning to
social participation,
the ``Common Vocabulary of Social Participation''
(\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social})
was proposed through a collaboration of Latin America academic, civil and governmental groups~\cite{corais}.
Started in 2012, by 2014
it already yielded relevant material, including a public preliminary \owl\ ontology with a concise taxonomy.
Also important are the reference documents reporting results from a first working phase, from July to December, 2012. 
As stated by the community, \vcps\ was propelled by three goals:
1) ease adoption of the vocabulary;
2) stimulate the creation of public tools to understand, visualize and summarize how social participation was being performed;
3) meet the needs of participatory initiatives to open and link their data. 
It is important to notice that \vcps,
an ontology, was called a \emph{vocabulary} both to ease understanding of the general public and because it started as a vocabulary.
The present article presents the ``Ontology of Social Participation'' 
(\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}),
based on \vcps, in which the term \emph{vocabulary} was substituted by the term \emph{ontology} for the following reasons:
\begin{itemize}
    \item the usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology
    (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \opsi\ is, in fact, an ontology, with a vocabulary,
    a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for academic acceptance and further formal adoption by government instances,
    such as the Brazilian Federal Portal of Open Data~\footnote{\url{http://dados.gov.br/}} and the Brazilian Federal Portal of Social Participation~\cite{participa}.
\end{itemize}

\noindent Also, the term \emph{common} was dropped when \ops\ was conceived,
as the term is redundant for an \owl\ ontology.
The \vcps\ presented other difficulties, such as missing classes,
incorrect \uri\ specifications (e.g. containing spaces), some logical flaws,
and unnecessary out-of-standards restrictions.
This were all solved within \ops.

Next subsection briefly presents most relevant related work.
Section~\ref{exp} presents \ops\ and modifications made from \vcps\ to \ops: class and property names and labels,
class restrictions and property axioms.
Section~\ref{ospUtil} is dedicated to \ops\ usage: dereferencing, \sparql\ queries,
a toy \ops\ expansion, a discursive example of usage, and use cases from government,
civil society and academic parties.
Concluding remarks are stated with future work, in Section~\ref{conc}. 
The Appendixes hold directions for the scripts that are used to obtain \ops,
notes on restrictions that were in \vcps\ but withdrawn from \ops\, and an inspection of the first-hand
documentation about \vcps\ (reference textual documents, images, \owl\ code, blog posts, discussions and \etherpad s).

\subsection{Related work}\label{srel}
There is a number of OWL ontologies for Social Network Systems (SNS, e.g.~\cite{sns1,sns2,sns3}),
although only few are dedicated to e-Participation (e.g.~\cite{ep1,ep2,ep3}).
Most importantly, the ontologies described in~\cite{pnud5} (\opa, \ocd, \ontologiaa, and \obs) were developed in compliance with \ops,
a fact further exploited in Section~\ref{ospUtil}.
The authors only found ontologies related to e-Participation, i.e. to social participation
performed in virtual platforms.
Thus, beyond the already cited constributions, authors considered suitable to mention some ontologies which may be related to social participation
as to enhance their utility.
Electronic Government (e-Gov) and Electronic Democracy (e-Democracy) ontologies are abundant (e.g.~\cite{ed1,ed2,ed3,ed4,eg1,eg2,eg3,eg4,eg5,eg6}), encompassing the expression of different institutions, processes and roles as linked data, and provide specific mechanisms for supporting social participation in the proposal, development and creation of public policies.
Other examples of ontologies promptly related to \ops\ are those related to human needs and smart cities (e.g.~\cite{sc1,hn}).
In summary, OPS is potentially unique in presenting a conceptualization about social participation in general
(i.e. not only as performed through online systems),
and in being a simple upper ontology that is the product of a collective effort.

% ontologies for social participation
% ontologies for social networks

% ontologias no PNUD

% 1) quais são as ontologias que já foram criadas para redes de participação. Vi que há alguns
% exemplos para redes sociais (como em algumas referências abaixo).
% 2) há alguma ontologia na literatura específica para portais como o participa.br?
% 3) do ponto de vista conceitual, qual é a inovação do trabalho?
% Notei que no artigo, não há referências recentes (2014 - 2019), e então será necessário
% verificar a literatura recente.
% abraço
% Chu
% 
% http://dare.ubvu.vu.nl/bitstream/handle/1871/33927/197060.pdf?sequence=1
% 
% https://link.springer.com/content/pdf/10.1007/978-3-540-72667-8_20.pdf
% 
% https://www.researchgate.net/publication/235407115_An_Ontology_for_Social_Networking_Sites_Interoperability
% 
% https://pdfs.semanticscholar.org/8759/9ad7df28cb5a528de36931548991c650aa75.pdf

\section{OPS: the Ontology of Social Participation}\label{exp}
\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (\ops).
    Arrows with white heads indicate ``is a'' relations (subclass points to superclass).
    Arrows with black heads indicate property relations from subject to object .
    All property relations yield existential restrictions,
    with the exception of the ``has characteristic'' property,
    that does not yield restriction.
    Upper ontologies \bfo\ and \foaf\ are asserted through color (\bfo) and dashed boxes (\foaf).}
    \label{fig:v1}
\end{figure*}
This section makes considerations about \ops\ label standardization and implemented classes,
properties and restrictions.
Features present in Figure~\ref{fig:v1}, but not present in \vcps,
are fully described in Section~\ref{impl}.
Usage examples are addressed in Section~\ref{ospUtil}.

\subsection{Standardization and implemented features}\label{impl}
Without explicit criteria,
\vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}.
\opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations.
    This is important to coalesce interested parties: the Brazilian Federal Social Participation Portal\cite{participa},
    the Brazilian government repository of vocabularies and ontologies~\cite{vocab},
    academic groups,  \ngo s, and non-organized civil society.
    \item Derived \uri s, when reached via \http, can be redirected to where current documentation is held,
    as it is hosted by \url{http://purl.org}.
\end{itemize}
Labels in the languages of interest should be written in label fields.
Even so, we advocate that \ops\ class names should be user-friendly,
bearing the attention not to take the class name as the label or as a meaning restriction.
For standardization, all classes are written in CamelCase~\cite{cc} in plain English to ease internationalization,
adoption and maturation.
Labels are written in English, Portuguese and Spanish.
Accordingly, class names changed from \vcps\ to \ops\,
received labels (\texttt{rdfs:label}) in each language
and a textual short explanation (\texttt{rdfs:comment}) in English. 
Table~\ref{ospClasses} presents all classes is \ops.
\opsi\ property names are written in the headlessCamelCase~\cite{cc} format,
are readable in English (to ease internationalization, adoption and maturation),
and some of them have defined domains and ranges.
Table~\ref{ospProps} is dedicated to \ops\ properties, with labels in English, Portuguese and Spanish.

In the first versions of \ops, all properties yielded existential restrictions,
except {\tt ops:hasParticipationCharacteristic}.
Although such efforts were aimed at providing a better defined \ops,
further developments and discussions revealed that these restrictions made \ops\ rigid,
a bit more complicated, and was not of much help,
at least in this stage of \ops\ development and adoption.
The result is that all restrictions were removed.
Table~\ref{ospRestr} (at Appendix~\ref{ap:restr}) presents all dropped restrictions from \vcps\ to \ops.

An inspection of \vcps\ reveals the contradiction found in Figure~\ref{fig:diaorig}:
a role cannot execute, initiate or support a cause.
These are performed by a social actor.
Therefore, to maintain \vcps\ directions and obtain a sound ontology,
{\tt ops:Role} was removed from \ops.
The result can be seen in Figures~\ref{fig:v1} and~\ref{fig:owlCC}.

@@

A comparison of the \vcps\ \owl\ code~\cite{owlCCPtg} 
with the diagram in Figure~\ref{fig:diaorig}, which reflects official \vcps\ documentation,
revealed that a class, two properties and three restrictions were not implemented in \vcps.
These were fully implemented in \ops\.
While restrictions missing in \vcps\ and implemented in preliminary \ops\ versions are presented in Appendix~\ref{ap:restr},
the missing classes and properties are:
\begin{itemize}
    \item Class: {\tt    ops:ParticipationCharacteristic}.
    \item Property: {\tt ops:hasRole}.
    \item Property: {\tt ops:composesSolution}.
\end{itemize}

\noindent \opsi\ is available online~\cite{owlOSP}.
To ease navigation of the ontology by interested parties,
it is also available in the Webprotege interface~\cite{owlOSPwp}.
A diagram of \ops ' taxonomic structure is provided in Figure~\ref{fig:owlCC}.

Upper ontologies usage with \ops\ is under development and should receive a dedicated article.
Pertinent and already used as upper ontologies for \ops\ are \foaf
(for ``linking and describing people and things they do''~\cite{foaf}) and \bfo
(``designed for use in supporting information retrieval, analysis and integration in scientific and other domains''~\cite{bfo}).
Upper ontologies classes related to each \ops\ class are also exposed in Table~\ref{ospClasses}.

Figure~\ref{fig:v1} is a complete diagram of current \ops:
classes, properties and relations to \foaf\ and \bfo.
Noteworthy, Figure~\ref{fig:v1} is more informative than the \ops\ \owl\ code,
as restrictions were removed and not all properties have defined domain and range.
Therefore, the diagram is an important source of relations envisioned by \ops\ creators.

@@

\section{OPS usefulness}\label{ospUtil}
\opsi\ is meant to be useful in at least two ways.
First, as a systematization of what is social participation to Latin America groups,
as conceived by \vcps.
Second, as way to facilitate the representation of social participation-related data as linked data,
and enable integration of various instances for social participation.
An indicative of this pertinence is \opa, \ocd, \ontologiaa, and \obs,
ontologies that already use \ops\ as upper ontology~\cite{pnud5}.

\opsi\ usage might not be obvious at first.
How is data linked? How is field knowledge organized? Why and by whom?
Core principles of \ops' utility can be understood by the following observations:
\begin{itemize}
    \item different participation instances have social actors,
    actions being performed, organizations involved, problems being tackled, etc.
    These can yield one consistent dataset by means of \ops\ usage.
    \item One can understand and share the mutually exclusive nature of being
    a paid or a voluntary contributor by observing the expanded version of \ops\ (see Section~\ref{downwards}).
    Also, noticing the fact that a mob can be very big or not,
    and that it can be convoked or not by a network,
    may make the field conceptualization more clear for a newcomer
    or ease discussions and problematization for senior researchers or politicians.
    \item Other fields of human knowledge and practice also have agents, problems and so forth.
    These can be linked to participatory data and ontologies by means of \ops.
    \item \opsi\ holds a conceptualization that is the product of the contribution of many specialists.
    In other words, put aside the linked data aspects,
    it is useful as an organization of the knowledge related to the social participation field that is representative of current understandings.
\end{itemize}
